 

### 2023-05-10 markdown/methods.Rmd
umap halfway explained

#### Batch correction 
Since mass spectrometry experiments are highly biased by several factors, such as sample processing, the technician, manufacturer and production of the device (chromatography column, mass-spectrometer), the reagents etc., working in multiple batches introduces further variance in the protein data. This variance would overshadow the biological differences across sample types. Before attempting to remove batch effects it is recommended to elucidate the variance with a dimensionality reduction technique which will be explained in the next chapter. To remove the technical factor also called non-biological experimental variation the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used. The expression of a gene $g$ for the batch $m_i$ and sample $j$ is defined as  $Y_i_j_g = a_g + X \beta_g + \gamma_i_g + \delta_i_g \epsilon_i_j_g$, whereas $\gamma$ and $\delta\epsilon$ denote for the systematic error caused by the batch. In order to remove the systematic error the bayesian framework shrinks the effect by pooling information over the features $g$. Ideally the effects $a$ and $\beta$ remain as natural as they are and should reflect the true biological difference between experimental obtained samples or groups. ComBat standardizes the data by a Z-transformation before attempting to remove the batch effect. By the method of moments the two batch effect parameters $\gamma$ and $\delta\epsilon$ were estimated and can be subtracted from each value for the particular feature.

#### Downstream Analysis
After cleaning the data and crystallizing from impurities such as noise and unwanted effects, the biological differences between sample types can be obtained. Proteomic data consists of multi dimensions (=features) which are overall hard to understand by a glance. An approach to observe the differences without a closer sight of exact effect on the proteinaceous ensemble, but still keep a high level of information can be dimensionality reduction.

##### Principle component analysis (=PCA)
The most common approach in data science for any kind of high dimensional data is the PCA. Every principle component can be understood as a vector pointing in a specific direction derived as the eigenvector from the covariance matrix of the expression set. The covariance matrix indicates for the co linearity between the variance of all elements compared. After calculating the covariance matrix the eigenvectors for the covariance matrix will be obtained by solving the quadratic equations to obtain the eigenvalues $\lambda$. Since $A v = \lambda v$ the eigenvector $v$ can be obtained by solving the equation for each element of the vector. The eigenvectors $v$ never changes is direction and explains the covariance within the dataset and so also the mathematical differences between the samples. The eigenvectors are sorted according to their eigenvalues in decreasing order and therefore decreasing explained variance. For graphical visualizations the principle components are plotted against each other. The first two components, which explain the majority of variance in the dateset highlight the difference between samples. After the cleaning procedures any additional effect caused by technical or biological replication should be removed, however it could be found be highlighting according to the particular factor. This makes the PCA to a useful tool before starting any statistical analysis by checking for involuntarily introduced bias.

>>>>

#### Batch correction 
Since mass spectrometry experiments are highly biased by several factors, such as sample processing, the technician, manufacturer and production of the device (chromatography column, mass-spectrometer), the reagents etc., working in multiple batches introduces further variance in the protein data. This variance would overshadow the biological differences across sample types. Before attempting to remove batch effects it is recommended to elucidate the variance with a dimensionality reduction technique which will be explained in the next chapter. To remove the technical factor also called non-biological experimental variation the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used. The expression of a gene $g$ for the batch $m_i$ and sample $j$ is defined as  $Y_{ijg} = a_{g} + X\beta_{g} + \gamma_{ig} + \delta_{ig}\epsilon_{ijg}$, whereas $\gamma$ and $\delta\epsilon$ denote for the systematic error caused by the batch. In order to remove the systematic error the bayesian framework shrinks the effect by pooling information over the features $g$. Ideally the effects $a$ and $\beta$ remain as natural as they are and should reflect the true biological difference between experimental obtained samples or groups. ComBat standardizes the data by a Z-transformation before attempting to remove the batch effect. By the method of moments the two batch effect parameters $\gamma$ and $\delta\epsilon$ were estimated and can be subtracted from each value for the particular feature.

#### Dimensionality reduction
After cleaning the data and crystallizing from impurities such as noise and unwanted effects, the biological differences between sample types can be obtained. Proteomic data consists of multi dimensions (=features) which are overall hard to understand by a glance. An approach to observe the differences without a closer sight of exact effect on the proteinaceous ensemble, but still keep a high level of information can be dimensionality reduction. After the cleaning procedures any additional effect caused by technical or biological replication should be removed, however it could be found be highlighting according to the particular factor. This makes the dimensionality reduction a useful tool before starting any statistical analysis by checking for involuntarily introduced bias.

##### Principle component analysis (PCA)
The most common approach in data science for any kind of high dimensional data is the PCA. For the pipeline the bioconductor package scater \citep{McCarthy2017} was used. Every principle component can be understood as a vector pointing in a specific direction derived as the eigenvector from the covariance matrix of the expression set. The covariance matrix indicates for the co linearity between the variance of all elements compared. After calculating the covariance matrix the eigenvectors for the covariance matrix will be obtained by solving the quadratic equations to obtain the eigenvalues $\lambda$. Since $A v = \lambda v$ the eigenvector $v$ can be obtained by solving the equation for each element of the vector. The eigenvectors $v$ never changes is direction and explains the covariance within the dataset and so also the mathematical differences between the samples. The eigenvectors are sorted according to their eigenvalues in decreasing order and therefore decreasing explained variance. For graphical visualizations the principle components are plotted against each other. The first two components, which explain the majority of variance in the dateset highlight the difference between samples.

##### Uniform manifold approximation & projection (UMAP)
UMAP \citep{McInnes2018} is a dimensionality reduction technique that was developed in recent years and is based on Riemann geometry. It shares similarities to the  t-distributed stochastic neighbor embedding (t-SNE) in terms of graphical visualization. Riemann geometry is a mathematical framework that deals with three dimensional functions using volumes, areas and vectors, which are also concept of topological data analysis. 
Before performing UMAP, three assumptions need to be considered:

\begin{enumerate}
\item There exists a manifold on which the data would be uniformly distributed.
$\rightarrow$ This assumption implies that the probability for each value in the biological dataset is the same.

\item The underlying manifold of interest is locally connected.
$\rightarrow$ In other words, similar samples will have a similar proteinaceous ensemble.

\item Preserving the topological structure of this manifold is the primary
goal.
$\rightarrow$ This means that the homeomorphic structure (the property of being able to transform one shape into another without tearing or gluing) is locally preserved, and local values in the dataset can be compared with each other.
\end{enumerate}
\citep{McInnes2018}

The UMAP algorithm begins by constructing a graph with a defined neighborhood parameter, typically denoted as $k$, which can be selected computationally. UMAP employs the k-nearest neighbor descent algorithm (NN-descent). Similar to the k-nearest neighbor algorithm (KNN), used in the missing value imputation part of this project, NN-descent recursively iterates to find the smallest distance to the respective neighbors within the defined neighborhood size $k$. Increasing the value of $k$ gradually strengthens the clustering of the data, but at a certain point, fine structure may be lost, and only rough estimates of the underlying principle can be observed.
The mathematical graph in UMAP is defined by nodes connected by edges, with each node internally connected to at least one other node. Edges are distinguished from each other not only by their connection but also by their importance, which is often referred to as weight.

The application utilizes the bioconductor package scater for performing UMAP \citep{McCarthy2017}.
